---
title: "beware_of_r-square"
format:
  html:
    df-print: kable
execute: 
  warning: false
---

## Setup environment

```{r}
#| output: false
library(tidyverse)
library(easystats)
library(ggformula)

set_theme(theme_bw())

set.seed(666)
```

## Overfitting

```{r}
N <- 10
x <- 1:N
alpha <- 20
beta <- 2
y <- alpha + beta * x + rnorm(N)
d <- tibble(x, y)
```

```{r}
gf_point(y ~ x, data = d)
```

```{r}
gf_point(y ~ x, data = d) |>
  gf_smooth(method = "lm", formula = y ~ 1, se = FALSE)
```

```{r}
gf_point(y ~ x, data = d) |>
  gf_smooth(method = "lm", formula = y ~ x, se = FALSE)
```

```{r}
gf_point(y ~ x, data = d) |>
  gf_smooth(method = "lm", formula = y ~ poly(x, N-1, raw=TRUE), se = FALSE)
```

```{r}
m0 <- lm(y ~ 1, data = d)
m1 <- lm(y ~ x, data = d)
m2 <- lm(y ~ poly(x, N-1, raw=TRUE), data = d)
```

```{r}
compare_performance(m0, m1, m2)
```

## Double descent

<https://www.r-bloggers.com/2025/10/double-descent-explained/>

```{r}
make_data <- function(n, sigma){
  x <- 2*(runif(n) - 0.5)
  y <- x * (x-0.6) * (x+0.6) + sigma * rnorm(n)
  data.frame(x=x, y=y)
}
```

```{r}
model_matrix <- function(dat, cutoffs, s_noise=0){
  n <- nrow(dat)
  N <- length(cutoffs)
  extra_features <- outer(dat$x, cutoffs, ">") + s_noise*rnorm(N*n)
  cbind(1, dat$x, extra_features)
}
```

```{r}
library(MASS)

fit_model <- function(dat, K, s_noise=0.01){
  # N is the number of random features added
  # dat is a univariate data set with columns x, y
  a <- min(dat$x)
  b <- max(dat$x)
  
  cutoffs <- runif(K)*(b-a) + a # random cutoffs

  bb <- ginv(model_matrix(dat, cutoffs, s_noise)) %*% dat$y
  list(bb=bb, cutoffs=cutoffs)
}

predict_model <- function(model, dat){
  # predict from the model from data set with column x
  model_matrix(dat, model$cutoffs, 0) %*% model$bb
}
```

```{r}
set.seed(2510)
dat <- make_data(50, 0.1)
model <- fit_model(dat, 20)
plot(dat, main="Training data and fitted model", las=1)
dat$pred <- predict_model(model, dat)
lines(dat$x[order(dat$x)], dat$pred[order(dat$x)], "l", col="red", lwd=2)
```

```{r}
calc_test_error <- function(model, nrow_test, sigma_test, n_reps){
  mse <- rep(0, n_reps)
  for (i in 1:n_reps){
    dat_test <- make_data(nrow_test, sigma_test)
    pred <- predict_model(model, dat_test)
    mse[i] <- mean((pred - dat_test$y)^2)
  }
  mean(mse)
}
```

```{r}
calc_variance <- function(n_pts, n_model, sigma, n_reps){
  x_test <- seq(-1, 1, len=n_pts)
  fitted <- matrix(0, nr=n_reps, nc=length(x_test))
  
  for (i in 1:n_reps){
    dat <- make_data(n_pts, sigma)
    model <- fit_model(dat, n_model)
    fitted[i,] <- predict_model(model, data.frame(x=x_test)) 
  }
  mean(apply(fitted, 2, var))
}
```

```{r}
#| cache: true
#| output: false
set.seed(2510)
K <- 50
model_reps <- 10
dat <- make_data(10, 0.1)
err <- vars <- matrix(0, nr=K+1, nc=model_reps)
for (i in 0:K){
  print(i)
  for (j in 1:model_reps){
    model <- fit_model(dat, i)
    err[i+1, j] <- calc_test_error(model, 100, 0.1, 1000)
    vars[i+1, j] <- calc_variance(10, i, 0.1, 1000)
  }
}
```

```{r}
plot_col <- rep("blue", K+1)
plot_col[1:9] <- "forestgreen" # used later
df <- data.frame(vars=rowMeans(log(vars)), 
                 err=rowMeans(log(err)), 
                 plot_col=plot_col)
plot(0:K, df$err, "l", xlab="Number of added features", ylab="log(test error)", las=1, main="Double Descent", col="blue", lwd=2)
abline(v=8, lty=2, col="red")
```

```{r}
plot(0:K, df$vars, "l", xlab="Number of added features", ylab="log(variance)", las=1, main="Model variance", lwd=2, col="darkgreen")
abline(v=8, lty=2, col="red")
```

```{r}
df$K <- 0:K
df <- df[order(df$vars), ]
plot(df$vars, df$err, col=df$plot_col, pch=19, xlab="log(variance)", ylab="log(test error)", las=1, cex=1.3)
loess_fit <- loess(df$err ~ df$vars, span = 0.6)
lines(df$vars, predict(loess_fit), col=rgb(0, 0, 0, 0.1), lwd=10)
```

## Double descent take two

```{r}
library(glmnet)
```

1. Simulate data: a simple sine wave with some noise

```{r}
set.seed(42)
n <- 100
x <- seq(-1, 1, length.out = n)
y <- sin(pi * x) + rnorm(n, sd = 0.2)
#y <- x + rnorm(n, sd = 0.2)
#y <- 1 + x + x^2 + x^3 + x^4 + rnorm(n, sd = 0.2)
```

2. Create high-order polynomial features

Using `raw=TRUE` to get standard powers: `x^1, x^2, ..., x^k`

```{r}
poly_degree <- 10000
X_poly <- poly(x, degree = poly_degree, raw = TRUE)
```

3. Define the order-dependent penalty weights

We penalize higher-order coefficients more heavily.

```{r}
weights <- log(seq_len(poly_degree))
```

4. Fit the model using `glmnet` with the `penalty.factor` argument

```{r}
fit <- cv.glmnet(X_poly, y, penalty.factor = weights)
```

5. Visualize the result

```{r}
y_pred <- predict(fit, newx = X_poly, s = "lambda.min")
plot(x, y, main = "Order-Dependent Regularization", pch = 19, col = "gray")
lines(x, y_pred, col = "blue", lwd = 2)
legend("topleft", legend = c("Data", "Model Fit"), col = c("gray", "blue"), pch = c(19, NA), lty = c(NA, 1))
```

## Print environment

```{r}
sessioninfo::session_info()
```
